---
title: "Out of Area Placements England"
author: "Vanessa Fillis"
date: "08/01/2021"
output: html_document
---

# Out of Area Placements in Mental Health Services in England

## Activate packages

```{r}
library(lubridate)
library(tidyr)
library(dplyr)
```

## Data import

In a previous step I have downloaded data for Out of Area Placements (OAPs) in mental health services between October 2016 and October 2020 (as this was the latest available data set) from [NHS Digital](https://digital.nhs.uk/data-and-information/publications/statistical/out-of-area-placements-in-mental-health-services). I have merged them, so that I only have one CSV to import. 

```{r}
OAPs <- read.csv("oaps2016to2021.csv", stringsAsFactors = FALSE)
```

## Inspect the data frame

What I can see:

- The column containing the date is not in date format, but as character. I have to add a date column that contains the publication period in date format. Because the data frame is large I will to this *after* I have filtered for the questions I want to look at, although this means I have to do it several times. But trying to do it before, led to R crashing. 
- The column **value** is in character format as well. I will transform it into numeric values. 

```{r}
head(OAPs)
OAPs$value <- as.numeric(OAPs$value)
head(OAPs)
```

## Filter for inappropriate OAPs in England that are active at the end of the period 

At first I want to look at inappropriate OAPs in England that are active at the end of each month. 

For almost all months the term **Out of Area Placement** is abbreviated. Only in July 2019, the full term has been used. This has to be considered when writing the filter. 

```{r}
OAPs.England.active <- subset(OAPs, OAPs$PublicationType == "Month")

OAPs.England.active <- subset(OAPs.England.active, OAPs.England.active$Question == "Inappropriate OAPs active at period end" | OAPs.England.active$Question == "Inappropriate out of area placements active at period end")

OAPs.England.active <- subset(OAPs.England.active, OAPs.England.active$Breakdown1 == "England")
```

### Convert date column

Now I will convert the date column. 

```{r}
#split date in PublicationPeriod in beginning and end of publication period
OAPs.England.active <- tidyr::separate(data = OAPs.England.active, col = PublicationPeriod, into = c("beginning.of.publication.period", "end.of.publication.period"), sep = "-")

OAPs.England.active$beginning.of.publication.period <- lubridate::dmy(OAPs.England.active$beginning.of.publication.period)

OAPs.England.active$end.of.publication.period <- lubridate::dmy(OAPs.England.active$end.of.publication.period)
```

### Delete empty columns

Some columns are empty. I will delete them. 

```{r}
OAPs.England.active <- OAPs.England.active[-c(7:11)]
```

### Export

Now I can export the data frame and will visualise it using Flourish. 

```{r}
write.csv(OAPs.England.active, file = "OAPs_Eng_active.csv")
```

## Filter for total inappropriate OAP days in England for each period

I also want to look at total inappropriate OAP days in England. The next steps will be very similar to the previous ones. 

```{r}
OAPs.England.days <- subset(OAPs, OAPs$PublicationType == "Month")

OAPs.England.days <- subset(OAPs.England.days, OAPs.England.days$Question == "Total number of inappropriate OAP days over the period" | OAPs.England.days$Question == "Total number of inappropriate out of area placement days over the period")

OAPs.England.days <- subset(OAPs.England.days, OAPs.England.days$Breakdown1 == "England")
```

### Convert date column

Now I will convert the date column. 

```{r}
#split date in PublicationPeriod in beginning and end of publication period
OAPs.England.days <- tidyr::separate(data = OAPs.England.days, col = PublicationPeriod, into = c("beginning.of.publication.period", "end.of.publication.period"), sep = "-")

OAPs.England.days$beginning.of.publication.period <- lubridate::dmy(OAPs.England.days$beginning.of.publication.period)

OAPs.England.days$end.of.publication.period <- lubridate::dmy(OAPs.England.days$end.of.publication.period)
```

### Delete empty columns

Some columns are empty. I will delete them. 

```{r}
OAPs.England.days <- OAPs.England.days[-c(7:11)]
```

### Export

Now I can export the data frame and will visualise it using Flourish. 

```{r}
write.csv(OAPs.England.days, file = "OAPs_Eng_days.csv")
```

## Filter for total inappropriate OAP costs in England for each period

I also want to look at costs caused by inappropriate OAPs each month. The next steps will be very similar to the previous ones. 

```{r}
OAPs.England.costs <- subset(OAPs, OAPs$PublicationType == "Month")

OAPs.England.costs <- subset(OAPs.England.costs, OAPs.England.costs$Question == "Total Cost For Inappropriate OAPs in Period" | OAPs.England.costs$Question == "Total Cost For Inappropriate out of area placements in Period")

OAPs.England.costs <- subset(OAPs.England.costs, OAPs.England.costs$Breakdown1 == "England")
```

### Convert date column

Now I will convert the date column. 

```{r}
#split date in PublicationPeriod in beginning and end of publication period
OAPs.England.costs <- tidyr::separate(data = OAPs.England.costs, col = PublicationPeriod, into = c("beginning.of.publication.period", "end.of.publication.period"), sep = "-")

OAPs.England.costs$beginning.of.publication.period <- lubridate::dmy(OAPs.England.costs$beginning.of.publication.period)

OAPs.England.costs$end.of.publication.period <- lubridate::dmy(OAPs.England.costs$end.of.publication.period)
```

### Delete empty columns

Some columns are empty. I will delete them. 

```{r}
OAPs.England.costs <- OAPs.England.costs[-c(7:11)]
```

### Export

Now I can export the data frame and will visualise it using Flourish. 

```{r}
write.csv(OAPs.England.costs, file = "OAPs_Eng_costs.csv")
```


## How many receiving providers are private in October 2020?

I want to look at where patients are being sent and if the receiving providers are private or NHS organisations. 

```{r}
OAPs.receivers <- subset(OAPs, OAPs$PublicationType == "Month")

OAPs.receivers <- subset(OAPs.receivers, OAPs.receivers$Question == "Inappropriate OAPs active at period end" | OAPs.receivers$Question == "Inappropriate out of area placements active at period end")

OAPs.receivers <- subset(OAPs.receivers, OAPs.receivers$Breakdown1 == "ReceivingProvider")
```

### Convert date column

```{r}
#split date in PublicationPeriod in beginning and end of publication period
OAPs.receivers <- tidyr::separate(data = OAPs.receivers, col = PublicationPeriod, into = c("beginning.of.publication.period", "end.of.publication.period"), sep = "-")

OAPs.receivers$beginning.of.publication.period <- lubridate::dmy(OAPs.receivers$beginning.of.publication.period)

OAPs.receivers$end.of.publication.period <- lubridate::dmy(OAPs.receivers$end.of.publication.period)
```

### Delete empty columns

```{r}
OAPs.receivers <- OAPs.receivers[-c(9:11)]
```

### Filter for October 2020

I want to look at the receiving organisations in October 2020. 

```{r}
OAPs.receivers.mar21 <- subset(OAPs.receivers, OAPs.receivers$beginning.of.publication.period == "2021-03-01")
```

### Using grepl()

To find out which of these organisations were private, I will look at their **Breakdown1Code**. 

I used [this table published by the NHS on organisation coding frames](https://datadictionary.nhs.uk/supporting_information/organisation_coding_frames.html) to find out what the codes stand for. 

- with 8, A, D, N = independent provider
-	with R, T = NHS trust or NHS Care trust

```{r}
OAPs.receivers.mar21$NHS <- grepl("^[R]|^[T]", OAPs.receivers.mar21$Breakdown1Code)
```

### Remove Unknown and England

Some receiving organisations are unknown. Additionally, the data frame also contains the total number of sending organisations. 

```{r}
OAPs.receivers.mar21 <- OAPs.receivers.mar21[!(OAPs.receivers.mar21$Breakdown1Code=="999" | OAPs.receivers.mar21$Breakdown1Code=="England"),]
```

### Remove those with 0 days

```{r}
OAPs.receivers.mar21 <- OAPs.receivers.mar21[!(OAPs.receivers.mar21$value=="0"),]
```

### Count true and false

```{r}
# How many receiving trusts are there overall?
sum(table(OAPs.receivers.mar21$NHS))

# How many are NHS (= TRUE) and how many private (= FALSE)?
table(OAPs.receivers.mar21$NHS)

## dividing the number of false values by the sum
74/95*100
```

### Export

```{r}
write.csv(OAPs.receivers.mar21, file = "OAPs_receivers_mar21.csv")
```


